---
title: "Getting started with pmotoolsr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pmotoolsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pmotoolsr` reads, writes, validates, processes, builds, and combines
**Portable Microhaplotype Objects (PMOs)** — the JSON format for targeted
amplicon (microhaplotype) sequencing data and its metadata. This vignette is a
quick tour using a small example PMO bundled with the package.

```{r setup}
library(pmotoolsr)

pmo_path <- system.file("extdata", "example_pmo.json.gz", package = "pmotoolsr")
pmo <- read_pmo(pmo_path)
```

## A note on indices: 0-based on disk, 1-based in R

A PMO links its sections with integer indices (a library sample points to its
specimen via `specimen_id`, a detected microhaplotype points into
`representative_microhaplotypes` via `mhaps_target_id`, and so on). **On disk
these indices are 0-based.** When read into R they are shifted to be **1-based**,
so they behave like ordinary R indices and can be used directly:

```{r}
lib <- pmo$library_sample_info[[1]]
pmo$specimen_info[[lib$specimen_id]]$specimen_name
```

On write, the shift is reversed automatically, so files stay interoperable with
[pmotools-python](https://github.com/PlasmoGenEpi/pmotools-python) and other
tooling. (`taxon_id` fields are not indices and are never shifted.)

Every `pmo_*` function accepts either a `read_pmo()` R6 object or a plain list
from `read_pmo_raw()`.

## Exploring a PMO

```{r}
length(pmo_get_specimen_names(pmo))
length(pmo_get_target_names(pmo))
pmo_get_panel_names(pmo)

head(pmo_count_targets_per_library_sample(pmo))
```

`pmo_has_section()` checks for optional sections before you use them:

```{r}
pmo_has_section(pmo, "sequencing_info")
```

## Validating

`pmo_validate()` runs structural checks plus full JSON Schema validation:

```{r}
pmo_validate(pmo)
```

## Subsetting

Filters return a new PMO (as a plain list) with every index remapped:

```{r}
some_targets <- pmo_get_target_names(pmo)[1:3]
sub <- pmo_filter_by_target_names(pmo, some_targets)
length(sub$target_info)

out <- tempfile(fileext = ".json.gz")
write_pmo_raw(sub, out)   # compression inferred from the extension
```

## Exporting tables

```{r}
head(pmo_export_specimen_meta_table(pmo))
head(pmo_extract_alleles_per_sample_table(pmo))
```

The allele table is the input many downstream tools (e.g. dcifer, moire) expect.
There are many more exporters (`pmo_export_*`), a BED extractor
(`pmo_extract_targets_insert_bed()`), and a multi-sheet Excel writer
(`pmo_export_to_excel()`).

## Where to next

- `vignette("building-a-minimal-pmo")` — build a PMO from a microhaplotype
  table and a primer panel, then add metadata.
- `vignette("building-a-full-pmo")` — assemble a fully-populated PMO from
  separate project, specimen, panel, library, sequencing, and bioinformatics
  tables.
```
